---
title: 'Gene set testing for Illumina HumanMethylation Arrays'
subtitle: "Exploring the biases on EPIC arrays "
author: "Jovana Maksimovic, Alicia Oshlack and Belinda Phipson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=10000)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
library(here)
library(glue)
library(minfi)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(missMethyl)
library(org.Hs.eg.db)
library(GO.db)
library(patchwork)
library(grid)
library(ggplot2)
library(tibble)
library(dplyr)
source(here("code/utility.R"))
```

## Array properties

Get the array annotation data.

```{r}
ann <- loadAnnotation(arrayType="EPIC")
```

Associate CpGs to genes (ENTREZ ID) using the Illumina annotation information.

```{r}
flatAnn <- loadFlatAnnotation(ann)
```

The number of CpGs annotated to a gene is highly variable.

```{r, fig.height=10, fig.width=8}
numCpgsPerGene <- as.vector(table(flatAnn$entrezid))
summary(numCpgsPerGene)
```

### Figure 1A {.tabset}
#### View

```{r}
dat <- data.frame(table(table(flatAnn$entrezid)))
numCpgsPerGene <- as.vector(table(flatAnn$entrezid))
med <- median(numCpgsPerGene)
mod <- getMode(numCpgsPerGene)

p <- ggplot(dat, aes(x=Var1, y=Freq)) +
    geom_segment(aes(x=Var1, xend=Var1, y=0, yend=Freq), color="darkgrey") +
    geom_point( color="black", size=1) +
    geom_vline(xintercept = med, linetype = "dashed", color = "red") +
    annotate("text", x = med + 2, label=glue("median = {med}"), y=720, colour="red", 
             size = 3, hjust="left") +
    geom_vline(xintercept = mod, linetype = "dashed", color = "red") +
    annotate("text", x = mod + 2, label=glue("mode = {mod}"), y=780, colour="red", 
             size = 3, hjust="left") +
    scale_x_discrete(breaks = c(1,10,20,30,45,55,70,80,90,100,120,135,150,160,180,
                                200,215,250,330,429,1485)) +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    xlab("No. CpGs per gene") +
    ylab("Frequency")

p
```

#### Create PDF

```{r, eval=FALSE}
fig <- here("output/Fig-1A.pdf")

pdf(file = fig, width = 9)
p
dev.off()
```

### Figure 1D {.tabset}
#### View
The number of genes a CpG maps to can also vary, although the majority of CpGs only map to one gene.

```{r}
dat <- data.frame(table(table(flatAnn$cpg)))
dat$Split <- ifelse(dat$Freq > 2500, "A", 
                    ifelse(dat$Freq < 2500 & dat$Freq > 65,"B","C"))

a <- ggplot(dat[dat$Split == "A",], aes(x=Var1, y=Freq)) +
    geom_bar(stat = "identity") +
    theme(axis.title.x = element_blank(), 
          axis.text.y=element_text(angle=90, hjust=0.5)) +
    ylab("Frequency")

b <- ggplot(dat[dat$Split == "B",], aes(x=Var1, y=Freq)) +
    geom_bar(stat = "identity") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank())

c <- ggplot(dat[dat$Split == "C",], aes(x=Var1, y=Freq)) +
    geom_bar(stat = "identity") +
    theme(axis.title.x=element_blank(),
          axis.title.y = element_blank()) 

p <- a + b + c + plot_layout(widths = c(1,2,6)) + 
    plot_annotation(caption = "No. genes mapped to a CpG",
                    theme = theme(plot.caption = element_text(hjust = 0.5, 
                                                              size = 12)))
p
```

#### Create PDF
```{r, eval=FALSE}
fig <- here("output/Fig-1D.pdf")
pdf(file = fig, width = 9)
p
dev.off()
```

### {-}

Explore the distribution of the average number of CpGs per gene, per GO category. 

### Figure 1C {.tabset}
#### View
```{r}
ann450 <- loadAnnotation("450k")
flatAnn450 <- loadFlatAnnotation(ann450)

cpgEgGoEPIC <- cpgsEgGoFreqs(flatAnn)
cpgEgGoEPIC$Array <- "EPIC"

cpgEgGo450 <- cpgsEgGoFreqs(flatAnn450)
cpgEgGo450$Array <- "450k"

cpgEgGoEPIC %>% bind_rows(cpgEgGo450) %>% 
    group_by(Array, GO) %>%
    summarise(avg = mean(Freq)) -> datSum

maxEPIC <- max(datSum$avg[datSum$Array == "EPIC"])
max450 <- max(datSum$avg[datSum$Array == "450k"])

bw <- 2
pal <- paletteer::paletteer_d("RColorBrewer::Set1", 2)
platform <- c("EPIC" = pal[1], "450k" = pal[2])
p <- ggplot(datSum, aes(x=avg)) +
    geom_histogram(dat = subset(datSum, Array == "450k"),
                   alpha = 0.5, aes(fill = "450k"), binwidth = bw) +
    geom_histogram(dat = subset(datSum, Array == "EPIC"),
                   alpha = 0.5, aes(fill = "EPIC"), binwidth = bw) +
    geom_vline(xintercept = maxEPIC, linetype="dashed", colour = pal[1]) +
    geom_vline(xintercept = max450, linetype="dashed", colour = pal[2]) +
    labs(x="Mean no. CpGs per gene per GO category", y = "Frequency",
         fill = "Platform") +
    annotate("text", x = maxEPIC - 4, label=glue("Max. EPIC = {maxEPIC}"),
             y = 1500, colour=pal[1], size = 2.5, hjust="right") +
    annotate("text", x = max450 - 4, label=glue("Max. 450k = {max450}"),
             y = 2500, colour=pal[2], size = 2.5, hjust="right") +
    scale_fill_manual(values = platform)
p
```

#### Create PDF
```{r, eval=FALSE}
fig <- here("output/Fig-1C.pdf")

pdf(file = fig, width = 9)
p
dev.off()
```

## Null simulations

Null simulation strategy: randomly select 50, 100, 500, 1000, 5000 and 10000 sets of CpGs and perform GO testing on each set 100 times, with and without adjusting for the various biases on the array. 

The code used for the simulation can be found in `r here("code/randCpgSim.R")`.   

### Figure 3A {.tabset}
#### View

The following boxplots show what proportion of the 100 simulations, at each level of CpGs sampled, had a raw p-value less than 0.05. This gives us an idea of the false discovery rate with and without adjustment for the number of CpGs annotated to a gene.

```{r}
dat <- readRDS(here("output/random-cpg-sims/EPIC.rds"))

dat %>% filter(method %in% names(dict)) %>%
    mutate(method = unname(dict[method])) %>%
    group_by(simNo, noCpgs, method) %>% 
    summarise(pSig = sum(P.DE < 0.05)/length(P.DE)) -> sigDat

p <- ggplot(sigDat, aes(x=noCpgs, y=pSig, fill=method)) + 
    geom_violin() +
    geom_hline(yintercept=0.05, linetype="dashed", color = "red") +
    labs(y="Prop. GO cat. with p-value < 0.05", x="No. randomly sampled CpGs",
         fill="Method") +
    scale_fill_manual(values = methodCols)
p
```

### Create PDF
```{r, eval=FALSE}
fig <- here("output/Fig-3A.pdf")

pdf(file = fig, width = 9)
p
dev.off()
```

## {-}

QQ plots of randomly selected simulations at each level of CpGs sampled.

```{r, fig.height=10, fig.width=7}
set.seed(42)
s <- sample(1:100, 3)
dat %>% filter(simNo %in% s) %>%
    filter(method %in% names(dict)) %>%
    mutate(method = unname(dict[method])) %>%
    arrange(simNo, noCpgs, method, P.DE) %>%
    group_by(simNo, noCpgs, method) %>%
    mutate(exp = 1:n()/n()) -> subDat

p <- ggplot(subDat, aes(x=-log10(exp), y=-log10(P.DE), color=method)) + 
    geom_point(shape = 1, size = 0.5) + 
    facet_grid(noCpgs ~ simNo, scales = "free_y") +
    scale_color_manual(values = methodCols)

p + geom_line(aes(x=-log10(exp), y=-log10(exp)),
              linetype="dashed", color = "black") +
    labs(y=expression(Observed~~-log[10](italic(p))), 
         x=expression(Expected~~-log[10](italic(p))),
         color="Method") +
    theme(legend.position="bottom", strip.text.x = element_blank())
```

### Figure 3A {.tabset}
#### View

Explore the relationship between the median, average number of CpGs, per gene, per GO category and the various sources of bias on the array.

```{r}
goFreq <- as_tibble(unique(cpgEgGoEPIC[,c("GO","Freq")]))

dat %>% filter(method %in% names(dict)) %>%
    mutate(method = unname(dict[method])) %>% 
    filter(P.DE < 0.05) %>%
    inner_join(goFreq, by=c("GO" = "GO")) %>%
    group_by(simNo, noCpgs, method, GO) %>%
    summarise(avgFreq=mean(Freq)) %>% 
    group_by(simNo, noCpgs, method) %>%
    summarise(medAvgFreq=median(avgFreq)) -> medAvgDat

p <- ggplot(medAvgDat, aes(x=noCpgs, y=medAvgFreq, fill=method)) + 
    geom_violin() + 
    stat_summary(geom="point", size=1, color="white", position = position_dodge(0.9),
                 show.legend = FALSE, fun = median) +
    labs(y="Median avg. no. CpGs/gene/GO cat.", 
         x="No. randomly sampled CpGs",
         fill="Method")  +
    scale_fill_manual(values = methodCols)
p
```

#### Create PDF
```{r, eval=FALSE}
fig <- here("output/Fig-3B.pdf")

pdf(file = fig, width = 9)
p
dev.off()
```

## {-}


