---
title: "Manuscript figures"
output: html_notebook
---

```{r}
library(here)
library(patchwork)
library(ggplot2)
library(limma)
library(tidyverse)
library(ggplotify)
library(UpSetR)
source(here("code/utility.R"))
options(scipen=10000)
```

The code in this document produces all of the panel figures in our manuscript 
and saves them as PDF files in the `output/figures` directory.

# Figures

## Figure 1

```{r}
a <- readRDS(here("output/figures/Fig-1A.rds"))
b <- readRDS(here("output/figures/Fig-1B.rds"))
c <- readRDS(here("output/figures/Fig-1C.rds"))
d <- readRDS(here("output/figures/Fig-1D.rds"))

layout <- "
AAAABB
CCDDDD
"
cairo_pdf(here("output/figures/Fig-1.pdf"), width = 9)
a + b+ c + theme(legend.position = "bottom") + 
    wrap_elements(panel = d, clip = FALSE) + 
    plot_layout(design = layout) + 
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Figure 3

```{r}
a <- readRDS(here("output/figures/Fig-3A.rds"))
b <- readRDS(here("output/figures/Fig-3B.rds"))
c <- readRDS(here("output/figures/Fig-3C.rds"))
d <- readRDS(here("output/figures/Fig-3D.rds"))

layout <- "
AABBBB
CCDDDD
"

b$layers[[4]]$aes_params$size <- 2.1

cairo_pdf(here("output/figures/Fig-3.pdf"), width = 9)
a + theme(legend.position = "none") + 
    b + theme(legend.position = "top",
              legend.key.size = unit(1, "lines")) +
    c + theme(legend.position = "bottom",
              legend.key.size = unit(1, "lines")) + 
    d + theme(legend.position = "bottom",
              legend.text = element_text(size = 8), 
              legend.key.size = unit(1, "lines"),
              axis.title.y = element_text(size = 9.5)) +
    plot_layout(design = layout) + 
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Figure 4

```{r}
a <- readRDS(here("output/figures/Fig-4A.rds"))
b <- readRDS(here("output/figures/Fig-4B.rds"))
c <- readRDS(here("output/figures/Fig-4C.rds"))
d <- readRDS(here("output/figures/Fig-4D.rds"))

layout <- "
AACCCC
AACCCC
BBCCCC
BBCCCC
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
"        
cairo_pdf(here("output/figures/Fig-4.pdf"), width = 8, height = 12)
a + theme(legend.position = "bottom", 
          legend.key.size = unit(0.75, "lines"),
          legend.text = element_text(size = 7),
          legend.title = element_text(size = 8),
          legend.box.just = 0,
          axis.title = element_text(size = 8),
          plot.margin = margin(0,0,0,0)) +
    b + theme(legend.position = "bottom", 
              legend.key.size = unit(0.75, "lines"),
              legend.text = element_text(size = 7),
              legend.title = element_text(size = 8),
              legend.box.just = 0,
              axis.title.y = element_text(size = 7),
              axis.text = element_text(size = 7),
              plot.margin = margin(0,0,0,0)) +
    c + guides(colour = guide_legend(ncol = 3)) +
    theme(legend.position = "bottom",
          legend.key.size = unit(1, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          plot.title = element_blank(),
          axis.title = element_text(size = 9)) +
    labs(y = "Prop. top 100 terms in truth set") +
    wrap_elements(d) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```
## Figure 5

```{r}
a <- readRDS(here("output/figures/Fig-5A.rds"))
b <- readRDS(here("output/figures/Fig-5B.rds"))
c <- readRDS(here("output/figures/Fig-5C.rds"))
d <- readRDS(here("output/figures/Fig-5D.rds"))
e <- readRDS(here("output/figures/Fig-5E.rds"))
f <- readRDS(here("output/figures/Fig-5F.rds"))
g <- readRDS(here("output/figures/Fig-5G.rds"))

layout <- "
AAACCCCC
AAACCCCC
BBBCCCCC
BBBCCCCC
DDDDDDDD
DDDDDDDD
DDDDDDDD
DDDDDDDD
DDDDDDDD
DDDDDDDD
EEEFFGGG
EEEFFGGG
EEEFFGGG
EEEFFGGG
EEEFFGGG
EEEFFGGG
"     

cairo_pdf(here("output/figures/Fig-5.pdf"), width = 9, height = 10)
a + theme(legend.position = "bottom", 
          legend.key.size = unit(0.75, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          legend.box.just = 0,
          axis.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          text = element_text(family = "arial")) +
    b + guides(fill = guide_legend(ncol = 3)) +
    theme(legend.position = "bottom",
          legend.key.size = unit(1, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          plot.title = element_blank(),
          axis.title = element_text(size = 8),
          text = element_text(family = "arial")) +
    c + theme(legend.position = "bottom", 
              legend.key.size = unit(1, "lines"),
              legend.text = element_text(size = 8),
              legend.title = element_text(size = 9),
              legend.box.just = 0,
              plot.title = element_blank(),
              axis.title = element_text(size = 8),
              text = element_text(family = "arial")) +
    wrap_elements(d + theme(plot.title = element_blank(),
                            legend.box = "horizontal",
                            legend.text = element_text(size = 7),
                            legend.key.size = unit(1, "lines"),
                            legend.title = element_text(size = 8),
                            axis.text.y = element_text(size = 7),
                            text = element_text(family = "arial"))) +
    wrap_elements(panel = e, clip = FALSE) +
    wrap_elements(panel = f + theme(axis.title = element_text(size = 9)), 
                  clip = FALSE) +
    wrap_elements(panel = g, clip = FALSE) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

# Supplementary Figures
## Supplementary Figure 1

```{r}
a <- readRDS(here("output/figures/SFig-1A.rds"))
b <- readRDS(here("output/figures/SFig-1B.rds"))

layout <- "
AAA
BBB
"
cairo_pdf(here("output/figures/SFig-01.pdf"))
a + wrap_elements(b) + 
    plot_layout(design = layout) + 
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 3

```{r}
a <- readRDS(here("output/figures/SFig-3A.rds"))
b <- readRDS(here("output/figures/SFig-3B.rds"))
c <- readRDS(here("output/figures/SFig-3C.rds"))
d <- readRDS(here("output/figures/SFig-3D.rds"))
e <- readRDS(here("output/figures/SFig-3E.rds"))

layout <- "
AABBBB
CCDDDD
EEEEEE
EEEEEE
EEEEEE
"

d$layers[[2]]$aes_params$size <- 0.2
a$layers[[1]]$aes_params$size <- 0.25

cairo_pdf(here("output/figures/SFig-03.pdf"), width = 8, height = 12)
a + theme(legend.position = "none", 
          axis.title = element_text(size = 9)) + 
    b + theme(axis.title.x = element_text(size = 9),
              axis.title.y = element_text(size = 8),
              legend.position = "top",
              legend.text = element_text(size = 8),
              legend.title = element_text(size = 9)) + 
    c + theme(axis.title = element_text(size = 9),
              legend.position = "bottom",
              legend.text = element_text(size = 8),
              legend.title = element_text(size = 9)) +
    d + theme(axis.title.x = element_text(size = 9),
              axis.title.y = element_text(size = 8),
              legend.position = "bottom",
              legend.text = element_text(size = 7),
              legend.title = element_text(size = 9, angle = 90)) +
    wrap_elements(e) +
    plot_layout(design = layout) + 
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 4

```{r}
a <- readRDS(here("output/figures/SFig-4A.rds"))
b <- readRDS(here("output/figures/SFig-4B.rds"))
c <- readRDS(here("output/figures/SFig-4C.rds"))
d <- readRDS(here("output/figures/SFig-4D.rds"))
e <- readRDS(here("output/figures/SFig-4E.rds"))
f <- readRDS(here("output/figures/SFig-4F.rds"))
g <- readRDS(here("output/figures/SFig-4G.rds"))

layout <- "
AAABBB
CCCDDD
EEEEEE
FFFGGG
"

cairo_pdf(here("output/figures/SFig-04.pdf"), width = 8, height = 10)
a + scale_y_continuous(labels = scales::scientific) +
    theme(legend.position = "right", 
          legend.box = "vertical",
          legend.text = element_text(size = 7),
          legend.title = element_text(size = 8),
          axis.title = element_text(size = 9),
          axis.text.x = element_text(size = 5)) +
    b + theme(legend.box = "vertical",
              legend.text = element_text(size = 7),
              legend.title = element_text(size = 8),
              axis.title = element_text(size = 9),
              axis.text.x = element_text(size = 5)) +
    c + theme(legend.text = element_text(size = 7),
              legend.title = element_text(size = 8)) +
    d + theme(legend.text = element_text(size = 8),
              legend.title = element_text(size = 8),
              axis.title = element_text(size = 8),
              axis.text = element_text(size = 7),
              legend.position = "bottom") +
    e + theme(legend.position = "bottom") +
    f +
    g + theme(legend.text = element_text(size = 8),
              legend.title = element_text(size = 8),
              axis.title = element_text(size = 8),
              legend.position = "right") +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```
## Supplementary Figure 5

```{r}
a <- readRDS(here("output/figures/SFig-5A.rds"))
b <- readRDS(here("output/figures/SFig-5B.rds"))

layout <- "
AAAA
BBBB
BBBB
BBBB
"        
cairo_pdf(here("output/figures/SFig-05.pdf"), width = 8, height = 9)
wrap_elements(a + theme(plot.title = element_blank(),
                        axis.title = element_text(size = 9),
                        legend.position = "right") +
                  labs(y = "Prop. top 100 terms in truth set")) +
    wrap_elements(b) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 6

```{r}
a <- readRDS(here("output/figures/SFig-6A.rds"))
b <- readRDS(here("output/figures/SFig-6B.rds"))

layout <- "
AAAA
BBBB
BBBB
BBBB
"        
cairo_pdf(here("output/figures/SFig-06.pdf"), width = 8, height = 9)
wrap_elements(a + theme(plot.title = element_blank(),
                        axis.title = element_text(size = 9),
                        legend.position = "right") +
                  labs(y = "Prop. top 100 terms in truth set")) +
    wrap_elements(b) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```
## Supplementary Figure 7

```{r}
a <- readRDS(here("output/figures/SFig-7A.rds"))
b <- readRDS(here("output/figures/SFig-7B.rds"))
c <- readRDS(here("output/figures/SFig-7C.rds"))
d <- readRDS(here("output/figures/SFig-7D.rds"))
e <- readRDS(here("output/figures/SFig-7E.rds"))

layout <- "
AAAA
AAAA
BBBB
BBBB
CCCC
CCCC
CCCC
CCCC
DDDD
DDDD
DDDD
DDDD
EEEE
EEEE
EEEE
EEEE
"        
cairo_pdf(here("output/figures/SFig-07.pdf"), width = 8, height = 22)
wrap_elements((a + plot_layout(guides = "collect") & 
                   theme(legend.box = "vertical", 
                         legend.position = "right"))) +
    wrap_elements((b + plot_layout(guides = "collect") & 
                       theme(legend.box = "vertical",
                             legend.position = "right"))) +
    c + 
    d +
    e +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 8

```{r}
f8a <- readRDS(here("output/figures/SFig-8A.rds"))
f8b <- readRDS(here("output/figures/SFig-8B.rds"))
f8c <- readRDS(here("output/figures/SFig-8C.rds"))
f8d <- readRDS(here("output/figures/SFig-8D.rds"))

layout <- "
AACCCC
AACCCC
BBCCCC
BBCCCC
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
DDDDDD
"        
cairo_pdf(here("output/figures/SFig-08.pdf"), width = 8, height = 12)
f8a + theme(legend.position = "bottom", 
            legend.key.size = unit(0.75, "lines"),
            legend.text = element_text(size = 7),
            legend.title = element_text(size = 8),
            legend.box.just = 0,
            axis.title = element_text(size = 8),
            plot.margin = margin(0,0,0,0)) +
    f8b + theme(legend.position = "bottom", 
                legend.key.size = unit(0.75, "lines"),
                legend.text = element_text(size = 7),
                legend.title = element_text(size = 8),
                legend.box.just = 0,
                axis.title.y = element_text(size = 7),
                axis.text = element_text(size = 7),
                plot.margin = margin(0,0,0,0)) +
    f8c + guides(colour = guide_legend(ncol = 3)) +
    theme(legend.position = "bottom",
          legend.key.size = unit(1, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          plot.title = element_blank(),
          axis.title = element_text(size = 9)) +
    labs(y = "Prop. top 100 terms in truth set") +
    wrap_elements(f8d) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 9

```{r}
a <- readRDS(here("output/figures/SFig-9A.rds"))
b <- readRDS(here("output/figures/SFig-9B.rds"))

layout <- "
AAAA
BBBB
BBBB
BBBB
"        
cairo_pdf(here("output/figures/SFig-09.pdf"), width = 8, height = 9)
wrap_elements(a + theme(plot.title = element_blank(),
                        axis.title = element_text(size = 9),
                        legend.position = "right") +
                  labs(y = "Prop. top 100 terms in truth set")) +
    wrap_elements(b) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 10

```{r}
a <- readRDS(here("output/figures/SFig-10A.rds"))
b <- readRDS(here("output/figures/SFig-10B.rds"))
c <- readRDS(here("output/figures/SFig-10C.rds"))
d <- readRDS(here("output/figures/SFig-10D.rds"))
e <- readRDS(here("output/figures/SFig-10E.rds"))
f <- readRDS(here("output/figures/SFig-10F.rds"))

layout <- "
AAABBBBB
CCCCCCCC
CCCCCCCC
DDDEEFFF
DDDEEFFF
"    

cairo_pdf(here("output/figures/SFig-10.pdf"), width = 8, height = 9)
a + guides(fill = guide_legend(ncol = 3)) +
    theme(legend.position = "bottom",
          legend.key.size = unit(1, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          plot.title = element_blank(),
          axis.title = element_text(size = 8)) +
    labs(y = "Prop. top 100 terms in truth set") +
    b + theme(legend.position = "bottom", 
              legend.key.size = unit(1, "lines"),
              legend.text = element_text(size = 8),
              legend.title = element_text(size = 9),
              legend.box.just = 0,
              plot.title = element_blank(),
              axis.title = element_text(size = 8)) +
    wrap_elements(c + theme(plot.title = element_blank(),
                            legend.box = "horizontal",
                            legend.text = element_text(size = 7),
                            legend.key.size = unit(1, "lines"),
                            legend.title = element_text(size = 8),
                            axis.text.y = element_text(size = 7))) +
    wrap_elements(panel = d, clip = FALSE) +
    wrap_elements(panel = e + theme(axis.title = element_text(size = 8)), 
                  clip = FALSE) +
    wrap_elements(panel = f, clip = FALSE) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 11

```{r}
a <- readRDS(here("output/figures/SFig-11A.rds"))
b <- readRDS(here("output/figures/SFig-11B.rds"))
c <- readRDS(here("output/figures/SFig-11C.rds"))
d <- readRDS(here("output/figures/SFig-11D.rds"))
e <- readRDS(here("output/figures/SFig-11E.rds"))
f <- readRDS(here("output/figures/SFig-11F.rds"))

layout <- "
AAABBBBB
CCCCCCCC
CCCCCCCC
DDDEEFFF
DDDEEFFF
"          
cairo_pdf(here("output/figures/SFig-11.pdf"), width = 8, height = 9)
a + guides(fill = guide_legend(ncol = 3)) +
    theme(legend.position = "bottom",
          legend.key.size = unit(1, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          plot.title = element_blank(),
          axis.title = element_text(size = 8)) +
    labs(y = "Prop. top 100 terms in truth set") +
    b + theme(legend.position = "bottom", 
              legend.key.size = unit(1, "lines"),
              legend.text = element_text(size = 8),
              legend.title = element_text(size = 9),
              legend.box.just = 0,
              plot.title = element_blank(),
              axis.title = element_text(size = 8)) +
    wrap_elements(c + theme(plot.title = element_blank(),
                            legend.box = "horizontal",
                            legend.text = element_text(size = 7),
                            legend.key.size = unit(1, "lines"),
                            legend.title = element_text(size = 8),
                            axis.text.y = element_text(size = 7))) +
    wrap_elements(panel = d, clip = FALSE) +
    wrap_elements(panel = e + theme(axis.title = element_text(size = 8)), 
                  clip = FALSE) +
    wrap_elements(panel = f, clip = FALSE) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 12

```{r}
a <- readRDS(here("output/figures/SFig-12A.rds"))
b <- readRDS(here("output/figures/SFig-12B.rds"))
c <- readRDS(here("output/figures/SFig-12C.rds"))
d <- readRDS(here("output/figures/SFig-12D.rds"))
e <- readRDS(here("output/figures/SFig-12E.rds"))

layout <- "
AAAA
BBBB
CCCC
CCCC
CCCC
CCCC
DDDD
DDDD
DDDD
DDDD
EEEE
EEEE
EEEE
EEEE
"        
cairo_pdf(here("output/figures/SFig-12.pdf"), width = 8, height = 17)
a + theme(legend.position = "right", 
          axis.title.y = element_text(size = 8)) +
    b + theme(legend.position = "none", 
              axis.title.y = element_text(size = 8)) +
    c + 
    d +
    e +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 13

```{r}
a <- readRDS(here("output/figures/SFig-13A.rds"))
b <- readRDS(here("output/figures/SFig-13B.rds"))

layout <- "
AAAA
BBBB
"        
cairo_pdf(here("output/figures/SFig-13.pdf"), width = 9)
a + b +
    plot_layout(design = layout, guides = "collect") +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 14

```{r}
a <- readRDS(here("output/figures/SFig-14A.rds"))
b <- readRDS(here("output/figures/SFig-14B.rds"))
c <- readRDS(here("output/figures/SFig-14C.rds"))
d <- readRDS(here("output/figures/SFig-14D.rds"))
e <- readRDS(here("output/figures/SFig-14E.rds"))
f <- readRDS(here("output/figures/SFig-14F.rds"))

layout <- "
AAABBBBB
CCCCCCCC
CCCCCCCC
DDDEEFFF
DDDEEFFF
"    

cairo_pdf(here("output/figures/SFig-14.pdf"), width = 8, height = 9)
a + theme(legend.position = "bottom", 
          legend.key.size = unit(0.75, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          legend.box.just = 0,
          axis.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          text = element_text(family = "arial")) +
    b + guides(fill = guide_legend(ncol = 3)) +
    theme(legend.position = "bottom",
          legend.key.size = unit(1, "lines"),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          plot.title = element_blank(),
          axis.title = element_text(size = 8),
          text = element_text(family = "arial")) +
    wrap_elements(c + theme(legend.position = "bottom", 
                            legend.key.size = unit(1, "lines"),
                            legend.text = element_text(size = 8),
                            legend.title = element_text(size = 9),
                            legend.box.just = 0,
                            legend.box = "horizontal",
                            plot.title = element_blank(),
                            axis.title = element_text(size = 8),
                            text = element_text(family = "arial"))) +
    wrap_elements(d + theme(plot.title = element_blank(),
                            legend.box = "horizontal",
                            legend.text = element_text(size = 7),
                            legend.key.size = unit(1, "lines"),
                            legend.title = element_text(size = 8),
                            axis.text.y = element_text(size = 7),
                            text = element_text(family = "arial"))) +
    wrap_elements(panel = e + theme(axis.text = element_text(size = 7),
                                    axis.title = element_text(size = 8)), 
                  clip = FALSE) +
    wrap_elements(panel = f + theme(axis.title = element_text(size = 9)), 
                  clip = FALSE) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## Supplementary Figure 15

```{r}
a <- readRDS(here("output/figures/SFig-15A.rds"))
b <- readRDS(here("output/figures/SFig-15B.rds"))
c <- readRDS(here("output/figures/SFig-15C.rds"))

layout <- "
AAAA
BBBB
BBBB
CCCC
"        
cairo_pdf(here("output/figures/SFig-15.pdf"), width = 8, height = 10)
wrap_elements(a + theme(legend.position = "right",
                        title = element_blank())) + 
    wrap_elements(shift_legend(b + theme(
        title = element_blank()))) + 
    wrap_elements(c + theme(title = element_blank())) +
    plot_layout(design = layout) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

