---
title: 'Gene set testing for Illumina HumanMethylation Arrays'
subtitle: "Restricting significant CpGs by genomic features"
author: "Jovana Maksimovic, Alicia Oshlack and Belinda Phipson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(here)
library(glue)
library(limma)
library(tidyverse)
library(patchwork)
if(packageVersion("missMethyl") != "1.21.5"){
    devtools::install_github("Oshlack/missMethyl", ref = "genomic-features")
}
library(missMethyl)
```

# Test restricting significant CpGs by genomic feature in gometh

Load methylation blood cell comparison input data and annotation.

```{r}
load(here("data/input.RData"))
anno <- minfi::getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
```

## Effect of restricting by genomic features on No. Sig. CpGs

For this dataset and these comparisons, restricting the significant CpGs by genomic feature reduces their numbers more dramatically than changing the FDR threshold from 0.05 to 0.01.

```{r, message=FALSE}
flatAnn <- missMethyl:::.getFlatAnnotation("EPIC", anno = anno)

dat <- NULL
for(i in 1:ncol(tfit$contrasts)){
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = 5000)),
                      cut = "Top 5000",
                      features = "ALL",
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    dat <- bind_rows(dat, tmp)
    
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = Inf, 
                                               p.value = 0.05)),
                      cut = "FDR < 0.05",
                      features = "ALL",
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    dat <- bind_rows(dat, tmp)
    
    feats <- c("TSS200","TSS1500","1stExon")
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = Inf, 
                                               p.value = 0.05)),
                      cut = "FDR < 0.05",
                      features = paste(feats,collapse = "."),
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    tmp %>% filter(cpgs %in% flatAnn$cpg[flatAnn$group %in% feats]) -> tmp
    dat <- bind_rows(dat, tmp)
    
    feats <- c("Body","ExonBnd")
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = Inf, 
                                               p.value = 0.05)),
                      cut = "FDR < 0.05",                      
                      features = paste(feats, collapse = "."),
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    tmp %>% filter(cpgs %in% flatAnn$cpg[flatAnn$group %in% feats]) -> tmp
    dat <- bind_rows(dat, tmp)
    
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = Inf, 
                                               p.value = 0.01)),
                      cut = "FDR < 0.01",
                      features = "ALL",
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    dat <- bind_rows(dat, tmp)
    
    feats <- c("TSS200","TSS1500","1stExon")
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = Inf, 
                                               p.value = 0.01)),
                      cut = "FDR < 0.01",
                      features = paste(feats,collapse = "."),
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    tmp %>% filter(cpgs %in% flatAnn$cpg[flatAnn$group %in% feats]) -> tmp
    dat <- bind_rows(dat, tmp)
    
    feats <- c("Body","ExonBnd")
    tmp <- data.frame(cpgs = rownames(topTreat(tfit, coef = i, num = Inf, 
                                               p.value = 0.01)),
                      cut = "FDR < 0.01",                      
                      features = paste(feats, collapse = "."),
                      contrast = colnames(tfit$contrasts)[i],
                      stringsAsFactors = FALSE)
    tmp %>% filter(cpgs %in% flatAnn$cpg[flatAnn$group %in% feats]) -> tmp
    dat <- bind_rows(dat, tmp)
}

dat %>% mutate(method = paste(features, cut, sep=".")) %>% 
    group_by(contrast, method) %>% tally() -> sub

ggplot(sub, aes(x = method, y = n, fill = method)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    facet_wrap(vars(contrast)) + 
    labs(fill = "Method", y = "No. significant CpGs", x = "Method") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Run gometh without and with restricting significant CpGs by genomic features. Examine imact on geneset testing results of restricting Cpgs significant at FDR < 0.05 to only certain genomic features. 

```{r}
outFile <- here("data/genomic-feature-go.rds")

if(!file.exists(outFile)){
    go <- NULL
    
    for(i in 1:ncol(tfit$contrasts)){
        top <- topTreat(tfit, coef = i, number = 5000)
        
        feats <- "ALL"
        tmp <- gometh(sig.cpg = rownames(top), 
                      all.cpg = rownames(tfit$coefficients), 
                      collection = "GO", array.type = "EPIC", anno = anno,
                      genomic.features = feats)
        tmp <- topGSA(tmp, number = Inf)
        tmp$FDR <- p.adjust(tmp$P.DE, method = "BH")
        tmp <- rownames_to_column(tmp, var = "ID")
        tmp$contrast <- colnames(tfit$contrasts)[i] 
        tmp$features <- paste(feats,collapse = ".")
        tmp$cut <- "Top 5000"
        go <- bind_rows(go, tmp)
        
        feats <- "ALL"
        tmp <- gometh(sig.cpg = rownames(topTreat(tfit, coef = i, number = Inf,
                                                  p.value = 0.05)), 
                      all.cpg = rownames(tfit$coefficients), 
                      collection = "GO", array.type = "EPIC", anno = anno,
                      genomic.features = feats)
        tmp <- topGSA(tmp, number = Inf)
        tmp$FDR <- p.adjust(tmp$P.DE, method = "BH")
        tmp <- rownames_to_column(tmp, var = "ID")
        tmp$contrast <- colnames(tfit$contrasts)[i] 
        tmp$features <- paste(feats,collapse = ".")
        tmp$cut <- "FDR < 0.05"
        go <- bind_rows(go, tmp)
        
        feats <- c("TSS200","TSS1500","1stExon")
        tmp <- gometh(sig.cpg = rownames(topTreat(tfit, coef = i, number = Inf,
                                                  p.value = 0.05)), 
                      all.cpg = rownames(tfit$coefficients), 
                      collection = "GO", array.type = "EPIC", anno = anno,
                      genomic.features = feats)
        tmp <- topGSA(tmp, number = Inf)
        tmp$FDR <- p.adjust(tmp$P.DE, method = "BH")
        tmp <- rownames_to_column(tmp, var = "ID")
        tmp$contrast <- colnames(tfit$contrasts)[i] 
        tmp$features <- paste(feats,collapse = ".")
        tmp$cut <- "FDR < 0.05"
        go <- bind_rows(go, tmp)
        
        feats <- c("Body","ExonBnd")
        tmp <- gometh(sig.cpg = rownames(topTreat(tfit, coef = i, number = Inf,
                                                  p.value = 0.05)), 
                      all.cpg = rownames(tfit$coefficients), 
                      collection = "GO", array.type = "EPIC", anno = anno,
                      genomic.features = feats)
        tmp <- topGSA(tmp, number = Inf)
        tmp$FDR <- p.adjust(tmp$P.DE, method = "BH")
        tmp <- rownames_to_column(tmp, var = "ID")
        tmp$contrast <- colnames(tfit$contrasts)[i] 
        tmp$features <- paste(feats,collapse = ".")
        tmp$cut <- "FDR < 0.05"
        go <- bind_rows(go, tmp)
        
    }
    saveRDS(go, file = outFile)
    
} else {
    go <- readRDS(outFile)
    
}
```

## Gene set testing results with/without restriction by genomic feature 

```{r}
immuneGO <- unique(read.csv(here("data/GO-immune-system-process.txt"), 
                            stringsAsFactors = FALSE, header = FALSE, 
                            col.names = "GOID"))

go %>% mutate(method = paste(cut, features, sep="; ")) %>% 
    arrange(contrast, method, P.DE) %>%
    group_by(contrast, method) %>%
    mutate(csum = cumsum(ID %in% immuneGO$GOID)) %>%
    mutate(rank = 1:n()) %>%
    filter(rank <= 100) -> dat

p <- ggplot(dat, aes(x = rank, y = csum, colour = method)) +
    geom_line() +
    facet_wrap(vars(contrast), ncol=3) +
    geom_vline(xintercept = 10, linetype = "dotted") +
    labs(colour = "Method", x = "Rank", y = "Cumulative no. immune sets") +
    theme(legend.position = "bottom", 
          legend.text = element_text(size = 8)) +
    guides(col = guide_legend(nrow = 2))
p
```

```{r}
immuneGO <- readRDS(here("data/RNAseq-GO.rds"))
immuneGO %>% group_by(contrast) %>%
    mutate(rank = 1:n()) %>%
    filter(rank <= 100) -> topSets

go %>% mutate(method = paste(cut, features, sep="; ")) %>%
    arrange(contrast, method, P.DE) %>%
    group_by(contrast, method) %>%
    mutate(csum = cumsum(ID %in% topSets$ID[topSets$contrast %in% contrast])) %>%
    mutate(rank = 1:n()) %>%
    filter(rank <= 100) -> sub

p <- ggplot(sub, aes(x = rank, y = csum, colour = method)) +
    geom_line() +
    facet_wrap(vars(contrast), ncol=3) +
    geom_vline(xintercept = 10, linetype = "dotted") +
    labs(colour = "Method", x = "Rank", 
         y = glue("Cumulative no. RNAseq sets")) +
    theme(legend.position = "bottom", 
          legend.text = element_text(size = 8)) +
    guides(col = guide_legend(nrow = 2))
p
```

Examine what the top 10 ranked gene sets are and how many genes they contain, for each approach and comparison.

```{r, fig.height=11, fig.width=6}
go %>% mutate(method = paste(cut, features, sep="; ")) %>% 
    arrange(contrast, method, P.DE) %>%
    group_by(contrast, method) %>%
    mutate(rank = 1:n()) %>%
    filter(rank <= 10) -> sub

p <- vector("list", length(unique(sub$contrast)) * length(unique(sub$method)))
i = 1
for(cont in unique(sub$contrast)){
    c = 1
    for(meth in unique(sub$method)){
        tmp <- sub %>% filter(contrast == cont & method == meth) %>%
            mutate(rank = factor(rank), 
                   rank = factor(rank, levels = rev(levels(rank))))
        
        p[[i]] <- ggplot(tmp, aes(x = -log10(FDR), y = rank)) + 
            geom_point(aes(size = N), alpha = 0.5, 
                colour = scales::hue_pal()(length(unique(sub$method)))[c]) +
            scale_y_discrete(labels = rev(tmp$TERM)) +
            labs(y = "", size = "No. genes", title = meth) +
            theme(axis.text.y = element_text(size = 6),
                  plot.title = element_text(size = 8),
                  legend.position = "right", 
                  legend.key.size = unit(0.25, "cm"),
                  legend.text = element_text(size = 6),
                  legend.title = element_text(size = 8),
                  axis.text.x = element_text(size = 6),
                  axis.title.x = element_text(size = 8)) + 
            coord_cartesian(xlim = c(-log10(0.99), -log10(10^-30))) +
            geom_vline(xintercept = -log10(0.05), linetype = "dashed")
        i = i + 1
        c = c + 1
    }
}

(p[[1]] / p[[2]] / p[[3]] / p[[4]]) + 
    plot_annotation(title = unique(sub$contrast)[1],
                    theme = theme(plot.title = element_text(size = 10))) 
(p[[5]] / p[[6]] / p[[7]] / p[[8]]) + 
    plot_annotation(title = unique(sub$contrast)[2],
                    theme = theme(plot.title = element_text(size = 10))) 
(p[[9]] / p[[10]] / p[[11]] / p[[12]]) + 
    plot_annotation(title = unique(sub$contrast)[3],
                    theme = theme(plot.title = element_text(size = 10))) 
```
